---
title: "regex-demo"
author: "Taavi Päll"
date: "4 10 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Find character strings

- Character strings can be matched and manipulated in base R by using regular expressions in functions grep, grepl, sub, gsub, regexpr + regmatches and some others. 

- tidyverse package ‘stringr’ contains analogous verbs with more consistent syntax. 

- A regular expression is a pattern that describes a set of strings.

## Regular Expressions as used in R

- Most characters, including all letters and digits, are regular expressions that match themselves. 

- Whereas, e.g. . matches any single character.

- You can refer also to a character class, which is a list of characters enclosed between [ and ], e.g. [[:alnum:]] is same as [A-z0-9].
- Most common character classes:
[:alnum:] includes alphanumerics ([:alpha:] and [:digit:]);
[:alpha:], includes alphabetic characters ([:upper:] and [:lower:] case);
[:punct:] includes punctuation characters ! " # $ % & ’ ( ) * + , - . / : ; < = > ? @ [  ] ^ _ ` ` { | } ~.;
[:blank:] includes space and tab; etc.

- The metacharacters in regular expressions are . \ | ( ) [ { ^ $ * + ?, whether these have a special meaning depending on the context.

- When matching any metacharacter as a regular character, precede it with a double backslash \\.

- Repetition quantifiers put after regex specify how many times regex is matched: ?, optional, at most once; *, zero or more times; +, one or more times; {n}, n times; {n,}, n or more times; {n,m}, n to m times.

- The caret ^ and the dollar sign $ are metacharacters that respectively match the empty string at the beginning and end of a line.

## Common operations with regular expressions

- Locate a pattern match (positions)

- Extract a matched pattern

- Identify a match to a pattern

- Replace a matched pattern

## Let's try out
Download test dataset.

> Test dataset contains Supplementary file names and some metadata of gene expresion profiling experiments using high-throughput sequencing: 

https://www.ncbi.nlm.nih.gov/gds?term=%22expression+profiling+by+high+throughput+sequencing%22[DataSet+Type]
```{r}
dir.create("data")
## manually download suppfilenames_2017-06-19.RData from rstats-tartu/datasets
```

## Load data
```{r}
load("data/suppfilenames_2017-06-19.RData")
```


## Unnest dataset
```{r}
library(tidyverse)
library(stringr)
supfn <- suppfilenames %>% 
  filter(!map_lgl(SuppFileNames, ~ inherits(., "try-error"))) %>% 
  unnest(SuppFileNames)
supfn %>% select(Accession, PDAT, SuppFileNames)
```


## Get string length
To get the length of a text string (i.e. the number of characters in the string).

```{r}
str_length("banana")
```


Length of supplementary file names.
```{r}

```

## Let's look at the filenames
```{r}
# Single most common filename: filelist.txt
most_common_filename <- supfn %>% 
  group_by(SuppFileNames) %>% 
  summarise(N=n()) %>% 
  arrange(desc(N))
most_common_filename
```

Filenames are prepended with GSE id 
```{r}
# Supplemental file names with more than N=10 occurences
cf <- supfn %>%
  mutate(common_filenames = str_replace(SuppFileNames, "GSE[0-9]*_", ""),
         common_filenames = str_replace(common_filenames, "\\.gz$", ""), 
         common_filenames = tolower(common_filenames))
```

```{r}
cfn <- group_by(cf, common_filenames) %>% 
  summarise(N=n()) %>% 
  arrange(desc(N)) %>% 
  filter(N > 10)

cfp <- ggplot(cfn, aes(common_filenames, N)) +
  geom_point() +
  scale_x_discrete(limits = rev(cfn$common_filenames)) +
  scale_y_log10() +
  coord_flip() + 
  xlab("Common stubs of SuppFileNames\n(>10 occurences) ") +
  ylab("Number of files")

# plot commonfilenames ggplot
cfp

```

```{r}
# Raw data as supplemental file plot (decide wether to show)
cfraw <- filter(cf, str_detect(common_filenames, "raw.tar")) %>%
  arrange(PDAT) %>%
  group_by(PDAT) %>%
  summarise(N = n()) %>%
  mutate(N = cumsum(N)) %>%
  ggplot(aes(ymd(PDAT), N, group = 1)) +
  geom_line()
```

